---
title: "Regex"
author: "Shannon Edie"
date: "January 28, 2020"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Assignment 4

First I read the dataset in using read.csv.

```{r}

MData <- read.csv("Data/measurements.csv")

```

Then, I used the dplyr package to create a new column for the dataset called "Sp", which contained only the first letter of the genus followed by a period and the species.

```{r}

library(dplyr)

MData <- mutate(MData, Sp= sub("\\B.+ ", ". ", Species))

library(knitr)
kable(MData, caption="\\label{mdata} Measurements data with both the species names and the abbreviated species names.")


```

I then created a table containing only the Sp column and not the Species column (I assumed that we were to keep all of the other columns, but if not please refer to the commented-out code).

```{r}

MData.2 <- select(MData, Sp, Limb.Width, unitsW, Limb.Length, unitsL)

kable(MData.2, caption="\\label{mdata.2} Measurements data with only the abbreviated species names.")
#MData.2 <- select(MData, Sp)

```

